Attrition in Online Learning: Understanding Persistence and Dropout in Massive Open Online Courses

### 2014-10-26 paper_v1.Rnw
corrected ROC AUC
added study 2 discussion

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey. One goal of this study was to further investiage research question $RQ3a$ based on learners' own descriptions of the challenges they faced. Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time. A variety of factors can influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments, are certainly large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:

\begin{description}
  \item[H4a] Learners who report having `not enough time' can be grouped into those with higher priority obligations (professional, academic, personal) and those with low volitional control.
  \item[H4b] Learners who do not report specific higher-priority obligations tend to exhibit lower volitional control than those who specify particular obligations.
\end{description}

Another goal of Study 2 was to collect data to test hypotheses $H1, H2, H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. To this end, we included a self-report measures for each construct in the survey that was sent out to at-risk learners.

@@

\textit{1. Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible. The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.

\textit{2. Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short. A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.

\textit{3. Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course. We thus had to train the prediction model using data from previous courses. The performance of the prediction model should be adequate in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the test set ($test_1$). Another second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model). The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the difference in ROC and AUC.

\begin{table}[h!]
\caption{Performance of the dropout prediction model}
\label{tab:drop_pred_rocs}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
ROC -- AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for a subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions\footnote{The $R$ mice package was used to perform multiple imputation and subsequent analyses.}. All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.

Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked. In addition, they reported intentions for engaging with each week's course content. Then, they were asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of. This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.

@@

Besides these open responses, learners rated their relative progress and satisfaction with their progress in the course, which served as measures of personal success. They also rated the extent to which their progress was hindered by a number of obstacles. The survey also included a measure of learners' sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and one item on volition (about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

The majority of survey respondents (73\%; 550 of 756) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. volition). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated, $r=.05, t_{548}=1.3 p=.21$.

Reasons for learner attrition were measured using open response as well as multiple-choice questions. Based on learners' coded open responses ($n=399$), 84\% had `not enough time for the course', 18\% indicated that `the course format or teaching style was not a good fit', 10\% expressed that `the level of the course was too advanced', 7\% found that `the course was not interesting or valuable enough', 4\% expressed a need for `additional course materials to learn the topic', 2\% experienced `technical difficulties', and 1\% reported facing a `language barrier'. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).

To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time, which revealed that about half of them (53\%) did not specify a particular committment. This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volition, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by committments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses 4a and 4b: while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volition and reporting no specific reasons for having not enough time was indicative thereof.

@@

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress. The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction. The results fully support $H1$, and partially support $H2$ and $H3$.

@@

Learners drop out at different times during the course. To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis. However, learners can develop risk factors at different times in the course. The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times. Such increased survey load can increase attrition or dissatisfaction in the course.

An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged. This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk). Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.

@@

>>>>

To gain a better understanding of the attrition patterns identified in Study 1, we designed a smaller, more focused follow-up study. In Study 2, we reached out to learners who were likely to drop out of a course and asked them to provide feedback. To this end, we developed a predictive model to identify at-risk learners and asked them to complete a survey. One goal of this study was to further investiage research question $RQ3a$ about reasons for attrition based on learners' own descriptions of the challenges they faced. Specifically, in Study 1, most learners reported that they disengaged due to a shortage of time. A variety of factors can influence the amount of time learners have at their disposal. Higher priority obligations, such as professional, academic, and personal committments, are certainly large contributors. Another critical factor, especially in the context of MOOCs as largely unguided learning environments, is learners' volitional control to engage in self-regulated learning \cite{corno2001volitional}. For instance, learners' ability to allocate time for the course in the presence of potential distractors of subjectively lower importance than taking the course (e.g., watching TV). We constructed the following hypotheses about the presence of these two factors for learners who report having not enough time:

\begin{description}
  \item[H4a] Learners who report having `not enough time' can be grouped into those with higher priority obligations (professional, academic, and personal) and those with low volitional control.
  \item[H4b] Learners who do not report specific higher-priority obligations tend to exhibit lower volitional control than those who specify particular obligations.
\end{description}

Another goal of Study 2 was to collect data to test hypotheses $H1, H2,$ and $H3$ on the role of goal striving, social belonging, and growth mindset in the success of MOOC learners. To this end, we included a self-report measures for each construct in the survey that was sent to at-risk learners.

@@

\textit{Prediction accuracy}: Prediction recall (the fraction of correctly predicted dropouts) should be high to deliver the survey to as many future dropouts as possible. The false positive rate (the fraction of non-dropouts incorrectly predicted as droputs) should be low to avoid sending a survey to learners who do not experience challenges in the courses.

\textit{Prediction lag}: The period between a learner's last site activity and the time they are flagged as being likely to drop out should be short. A shorter prediction lag improves the likelihood of successfully intervening to help the learner get back on track.

\textit{Transferability}: Whether or not a learner drops out of the course can not be determined before the end of the course. We thus had to train the prediction model using data from previous courses. The performance of the prediction model should be adequate in courses other than those on which it was trained.

The training set was composed of learners' interaction data (predictors) and dropout states (outcome) from 20 MOOCs. The large number of courses should improve model transferability by favoring features that are strongly predictive across different courses. Five hundred learners from each MOOC were included in the training set ($train$), and 500 other learners from the same MOOCs were selected for the first test set ($test_1$). A second test set of similar size was composed from 20 other MOOCs ($test_2$). Features extracted from the interaction data included different aspects of learners' video, assignment, and forum activity; their pace of viewing videos relative to the course pace, as well as various aspects of engagement time, such as the last time the learner interacted with the course (see \cite{halawa2014dropout}, for additional details on the prediction model). The model was fit by logistic regression. Table \ref{tab:drop_pred_rocs} provides information on the prediction performance, measured by the area under the ROC curve (AUC).

\begin{table}[h!]
\caption{Performance of the dropout prediction model}
\label{tab:drop_pred_rocs}
\small
\center
\begin{tabular}{lccc}
\toprule
 & $train$ & $test_1$ & $test_2$ \\
\midrule
AUC &  0.931 & 0.929 & 0.922  \\
\bottomrule
\end{tabular}
\end{table}

@@

Every learner who was predicted to disengage from the course was sent an email kindly requesting their help: ``You are enrolled in [course name], but you've been less active recently. Could you help us understand why?'' A low response rate was expected for this subpopulation that was defined by low engagement. And yet, 756 out of 6,053 learners started the survey (12.5\% response rate), and 459 completed it (61\% completion rate). Missing values were multiply imputated by predictive mean matching using responses to all survey questions.\footnote{The $R$ mice package was used to perform multiple imputation and subsequent pooled analyses.} All estimates were pooled across ten imputations, with standard errors adjusted for added variance. The majority of respondents were male (85\%), 35.6 years old on average ($SD=12.5$), and 80\% had achieved a bachelors' or more advanced degree.

Learners were asked to report how satisfied they were with their progress in the course, and whether they were using the course materials more, less, or exactly as much as they would have liked. These items served as measures of personal success.
% In addition, they reported intentions for engaging with each week's course content. 
They were then asked to openly report ``what challenges, inside or outside of the course, [they] experienced while [they were] taking this course, if any?'' The instructions encouraged them to list all challenges they could think of. This question was deliberately asked prior to any survey questions that could suggest particular reasons for disengagement.

@@

% Learners also rated the extent to which their progress was hindered by a number of obstacles. 
Learners also rated their sense of social and academic fit \cite{walton2007question} (17 items, $M=4.66, SD=0.71, \alpha=0.86$); mindset (4 items on the nature of intelligence and talent\footnote{Items were adpated from the mindset questionnaire available at \url{http://mindsetonline.com/testyourmindset/}.}; $M=4.44, SD=0.92, \alpha=0.77$), goal striving (4 items on motivation, perceived importance, committment, and confidence; $M=3.12, SD=0.95, \alpha=0.82$), and capacity for volitional control (one item about lower-priority distractions hindering the learner's course progress; $M=2.73, SD=1.29$).

@@

The majority of survey respondents (550 of 756; 73\%) reported being only somewhat or less satisfied with their progress in the course. While many of them (66\%) were substantially held back by other committments that took up time they had planned to spend on the course, only one in four (25\%) indicated being hindered by distractions of lower priority than the course (i.e. exhibit lower volitional control). These two time-related measures, which only differ in the assigned level of priority, were not significantly correlated ($r=.05, t_{548}=1.3, p=.21$).

Reasons for learner attrition were measured using iteratively coded open responses ($n=399$). Table \ref{tab:s2reas} provides each coded challenge and percentage of open responses that indicated it. A number of significant correlations between these challenges stood out: having not enough time was negatively associated with disliking the format or teaching style ($r=-.52$), needing additional course materials ($r=-.26$), finding the course uninteresting or not valuable ($r=-.45$), and facing technical difficulties ($r=-.29$).

\begin{table}[h!]
\caption{Iteratively-Coded Reasons for Attrition}
\label{tab:s2reas}
\small
\center
\begin{tabular}{lr}
\toprule
What challenges have you experienced while taking this course?  & \% \\
\midrule
Not enough time for the course & 84 \\
The course format or teaching style is not a good fit & 18 \\
The level of the course is too advanced & 10 \\
The course is not interesting or valuable enough & 7 \\
Additional course materials needed to learn the topic & 4 \\
Technical difficulties & 2 \\
Language barrier & 1 \\
\bottomrule
\end{tabular}
\end{table}

To test hypotheses $H4a$ and $H4b$, we examined the 336 open responses that indicated having too little time. It revealed that no particular commitment was specified in about half of them (53\%). This distinction was a salient feature in the initial codebooks, but could not be coded with sufficient reliability across coders. Coders would have required additional training to avoid reading specific reasons into unspecific resposnes. We thus operationalized the distinction by string matching to caputre professional, academic, and personal reasons for having not enough time.\footnote{Pattern: {\em work$|$job$|$school$|$university$|$college$|$family$|$kids$|$health}} Relative to those who specified a reason, learners who indicated too little time without specifying a reason also reported a significantly lower level of volitional control, i.e. their progress was more severely hindered by lower-prioirty distractions ($t_{331}=2.9, p=.003, d=.32$). However, the two groups did not significantly differ in how much they were held back by commitments that took up time they had planned to spend on the course ($t_{333}=-.75, p=.45$). These findings lend strong support to hypotheses $H4a$ and $H4b$: that is, while the large group of learners who report having not enough time generally faced higher-priority obligations, a substantial proportion suffered from low volitional control---and reporting no specific reasons for having not enough time was indicative thereof.

@@

Hypotheses $H1$, $H2$, and $H3$ about goal striving, perceived social belonging, and mindset, respectively, were tested by comparing self-described successful with unsuccessful learners. Success was measured by relative progress, and satisfaction with progress: 68\% of respondents reported using the course materials less than they would have wanted, and 57\% reported not being satisfied with their progress in the course (18\% were very or extremely dissatisfied). The two measures of success were correlated, $r=.35, t_{754}=10, p<.001$; satisfied respondents were, however, equally likely to be successful and unsuccessful in terms of their progress. Table \ref{tab:psych} provides summary statistics for each psychological measure by these two success metrics. Successful learners exhibited higher levels of goal striving according to both progress and satisfaction measures. Social belonging was significantly higher for learners who were successful in terms of satisfaction but not progress. The pattern was reversed for growth mindset, with evidence for the hypothesized effect for progress but not satisfaction. Hence, the results fully support hypothesis $H1$, and partially support $H2$ and $H3$.

@@

Study 2 focused on reasons for attrition and psychological factors of learners who were likely to drop out of the course. Through iterative coding, we developed a set of challenges that were faced by a subset of learners in a particular MOOC (cf. Table \ref{tab:s2reas}). The large majority of learners who indicate that they had not have enough time was found to be sperable into two groups: those with higher priority obligations that take up too much time, and those who are challenged by low levels of volitional control ($H4a$). MOOCs are an environment that requires learners to be highly self-regulated to be successful. The present findings emphasize the critical role of their capacity for volitional control, a factor that warrants further investigation \cite{corno2001volitional}. Very few learners explicitly indicated issues with volitional control in their responses, potentially because it is not socially desirable to admit. The absence of specific reasons for their shortage of time was found indicative of issues with volitional control ($H4b$). This could be a useful proxy for volitional issues in future research.

We investigated the extent to which personal success in MOOCs is associated with three established psychological factors: goal striving, social belonging, and growth mindset. Prior evidence found the absence of each factor to cause lower persistence and performance in more traditional achievement-oriented environments. The present findings highlight the critical role of these factors in MOOCs based on observational evidence. Learners with higher levels of goal striving, social belonging, and growth mindset also reported higher levels of personal success ($H1,H2,H3$). The subpopulation of learners in this study was a small and intentionally skewed subset. The observed differences are expected to be larger in the full population of learners, compared to this homogenous sample of learners who are unlikely to persist. Goal striving was most strongly associated with both measures of personal success. Success in terms of relative progress was associated with a stronger growth mindset, but not significantly with social belonging. In contrast, success in terms of satisfaction with progress was associated with stronger social belonging, but not significantly with growth mindset. Future work should examine if this pattern reflects a difference in the underlying process by which these psychological factors affect learner behavior and perception.

The methodology used in this study can be extended to deliver timely and targeted interventions. The predictive model was successful in identifying a large number of leaners who were facing challenges with taking the course. The majority of them were unsuccessful in their own terms. All three psychological factors under investigation were found to distinguish self-ascribed successful from unsuccessful learners. Recent work on psychological interventions found lasting positive effects of brief interventions involving reading and writing tasks (e.g., \cite{walton2007question}).  Leveraging predictive models of persistence, brief interventions could be delivered online and targeted to learners who face particular challenges, such as low goal striving, a fixed mindset, or perceived threat to their identity or social belonging.

@@

% Learners drop out at different times during the course. To elicit feedback from likely dropouts, one possible approach is to send the follow-up survey to every learner in the course, and then exclude learners who persist from the analysis. However, learners can develop risk factors at different times in the course. The drawback of this approach is that it requires the survey to be delivered to every learner multiple times in order to capture feedback from learners who might develop risk factors at different times. Such increased survey load can increase attrition or dissatisfaction in the course.

% An alternative approach is to employ a predictive model that red-flags a learner when they start exhibiting behaviors predictive of dropout, and sending the survey only to students who are newly red-flagged. This reduces survey burden while allowing us, ideally, to reach the whole population of interest (every learner who is at risk). Several MOOC instructors required minimization of the workload associated with taking the follow-up survey, which urged us to follow this approach.
